---
title: "Workshop: Introduction to redlistr"
author: "Aniko B. Toth"
format: html
editor: visual
---

## Overview - Lesson plan

A beginner workshop for using the `{redlistr}` package in **R** to calculate key spatial metrics used in the **IUCN Red Lists of Ecosystems and Species** workflow from ecosystem distribution rasters and other data formats.

**Lesson 1: Setup**

a.  Introduction to `redlistr` repository, R, RStudio, and Quarto

b.  Setting up the environment

c.  Importing data

d.  Checking data

e.  Visualising data

**Lesson 2: Extent of Occurrence (EOO)**

**Lesson 3: Area of Occupancy (AOO)**

**Lesson 4: Calculating Area and Rate of change**

## Lesson 1. Setup

### a. Introduction to `redlistr` repository, R, RStudio, and Quarto

We want to keep all of our files in one folder. Download a copy of the [GitHub repository](https://github.com/red-list-ecosystem/redlistr_intro_workshop.git) (use the green Code button at the top right), and save it in a convenient place on your computer. The repository will be your main workspace for this workshop. Open the workshop in Rstudio using the `redlistr_intro_workshop.Rproj` file. This will automatically set your working directory to the repository folder.

``` text
redlistr_intro_workshop/
├── Data
    ├── WhiteAshForest
        ├── (spatial files)
    ├── Plant_species_points.csv
├── Exersises
    ├── Exercises.qmd
    ├── Exercises_ANSWERS.qmd
├── .gitignore
├── redlistr_intro_workshop.proj
```

Before starting, install:

-   R (language)
-   RStudio (IDE - Integrated Development Environment)
-   Quarto (optional, but useful for workshop documents like this one)
-   Download the repository as zip and open the Rproj
-   Check out `README.md` for installing details and troubleshoots

### b. Setting up environment

#### Install Packages

Run this once on your computer if the packages are not already installed:

```{r}
#| eval: false
# Install the packages needed for this workshop.
# Run this only once on your computer, not every session.
install.packages(c(
  "sf",              # shapefile handling
  "terra",           # raster handling
  "leaflet",         # interactive maps
  ))

### before package is loaded to CRAN: 
install.packages("remotes")
remotes::install_github("red-list-ecosystem/redlistr", force = TRUE)
# if at this point you are prompted to update packages, enter 1 to update ALL and click "yes" in the popup window.
```

#### Load Packages

Run this each session when you start your analysis:

```{r Loading packagesmessage=FALSE}
# Load packages

library(sf)  # work with point, line, and polygon data
library(terra)  # work with raster data
library(leaflet) # build interactive maps
library(redlistr) # Red List of Ecosystems calculations
```

### c. Importing Data

#### Raster data

In this step, we first get the data, then visualise them and check if our data matches the metadata.

Here we will use the mangrove rasters bundled inside `{redlistr}`. Our data-set is based on Mangroves distributions around French Island VIC.

```{r Loading our example distributions}
# Read the example mangrove raster for year 2000 from the redlistr package.
mangrove.2000 <- rast(
  system.file("extdata", "example_distribution_2000.tif", package = "redlistr")
)

# Read the example mangrove raster for year 2017 from the redlistr package.
mangrove.2017 <- rast(
  system.file("extdata", "example_distribution_2017.tif", package = "redlistr")
)
```

-   Here, `system.file()` points to files bundled inside the package.

-   **When using your own data, replace this with a full file path, for example:** `my_raster <- rast("path/to/my_ecosystem_2000.tif")`

#### Species point data

Let's do one more example to address the import of point data (e.g., species occurrences). We will use this dataset later to demonstrate calculations with multiple assessment units at once.

```{r}
#Load the "Plant_species_points.csv" dataset from the Data folder in the work directory.
plant_species <- read.csv("../Data/Plant_species_points.csv") |> 

  # convert the table into a spatial points object by specifying the column names containing coordinates
  st_as_sf(coords = c("lon", "lat")) |>
  
  # specify the Coordinate system the data was provided in
    st_set_crs(4326) 

# Look at the resulting spatial points dataset
plant_species
#class(plant_species)
```

**⚠️ New R skill**

-   You will notice we have used the special operator `|>` in the code above. This is called a "pipe" and allows you to enter instructions sequentially, without saving intermediate products. A pipe takes the product of the code before it and inserts it into the first argument slot of the function after it. So, the code below has the same effect, but is much messier!
-   This snippet below calls a CSV file named `Plant_species_points.csv` (located in a `Data` folder one level up from your current directory), which contains plant species records with longitude and latitude columns. The `st_as_sf()` function converts it from a plain table into a **spatial points object**, telling R which columns hold the coordinates (`lon` and `lat`). (**Spatial CSVs**)

```{r}

intermediate1 <- st_as_sf(read.csv("../Data/Plant_species_points.csv"), coords = c("lon", "lat"))
plant_species <- st_set_crs(intermediate1, 4326)
plant_species
class(plant_species)
```

```{r}
# plain aspatial
aspatial <- read.csv("../Data/Plant_species_points.csv")
class(aspatial)       # "data.frame"
ncol(aspatial)        # 8 columns (including lon and lat)
names(aspatial)  # geometry column replaces lon + lat

# spatial
class(plant_species)  # "sf" "data.frame"
ncol(plant_species)   # 7 columns (lon & lat merged into geometry)
names(plant_species)  # geometry column replaces lon + lat
```

##### !!! For Alternative formats

See vignette [Ecosystem distribution data in raster and vector format](redlistr-raster-vector.html) for instruction on how to read ecosystem distribution data in other formats like shapefiles.

### b. Checking Data

It is critical to perform two checks **before any calculations**.

The `{redlistr}` functions require distances in **metres**, not degrees.

Your data must:

-   be in a **projected CRS** rather than geographic longitude/latitude

-   use the **same CRS** if comparing multiple times

Before doing any calculations, always check the Coordinate Reference System (CRS). The package will check this internally and throw an error if datasets are provided with geographic coordinates.

**Definition:** CRS (Coordinate Reference System) defines how spatial data are positioned on the Earth (e.g., latitude/longitude -- geodesic coordinate systems that model a round surface or projected coordinate systems where coordinates are projected onto a flat surface).

**Definition:** Projected CRS here represents data in metres (or feet), which is required for accurate area and distance calculations, as degrees can represent different distances depending on latitude.

Note: For Red List assessments, it is best to choose a CRS that compromises between accurate areas, distances, and bearings. For smaller extents, the exact CRS will not affect the results, but for analyses covering larger, regional areas, we might like to test if different projections yield different results.

##### Let's check CRS of **both** rasters

```{r}
# Print CRS of 2000 raster
st_crs(mangrove.2000)

# Print CRS of 2017 raster
st_crs(mangrove.2017)

# Example: choose a projected CRS suitable for your study area
# mangrove.2000 <- project(mangrove.2000, "EPSG:32754")
# mangrove.2017 <- project(mangrove.2017, "EPSG:32754")

# transform points to a projected CRS (UTM)
plant_species <- st_transform(plant_species, 3577)
plant_species
```

#### Check if data are in longitude/latitude (degrees)

```{r}
# TRUE = geographic (degrees), FALSE = projected (metres)
# terra function for rasters
is.lonlat(mangrove.2000)
is.lonlat(mangrove.2017)

# sf function for shapefiles
st_is_longlat(plant_species)


```

If `is.lonlat()` or `st_is_longlat()` returns TRUE, your data are in degrees (longitude/latitude).

You must reproject them before continuing. We will not cover selecting an appropriate CRS in this workshop, but external resources for this step are available. The `rast` package contains the `project()` function for reprojecting rasters, while the `sf` function `st_transform()` is used for points, lines, or polygons. Note the differing notation for specifying the CRS.

#### Compare CRS between layers

```{r}
# Check if both rasters use the same CRS
crs(mangrove.2000) == crs(mangrove.2017)
```

### c. Visualising data

Before doing any calculations, inspect what the rasters look like.

`add = TRUE` overlays the 2017 map on top of the 2000 map.

-   **Grey** = 2000

-   **Orange** = 2017 `rast()` specifies that the file is to be read in as a raster file.

Therefore, areas that show up in grey are those where mangrove distribution has been lost between 2000 and 2017.

```{r Plotting the two rasters, fig.show='hold', fig.width=7, fig.height=7, fig.alt='Overlay of two mangrove distribution maps'}
plot(
  mangrove.2000,        # Base layer: mangrove distribution in year 2000
  col = "grey30",       # Grey colour for 2000 presence
  legend = FALSE,       # Remove legend for cleaner overlay
  main = "Mangrove Distribution"  # Plot title
)

plot(
  mangrove.2017,        # Overlay layer: mangrove distribution in 2017
  add = TRUE,           # Add on top of existing plot (do not create new plot)
  col = "darkorange",   # Orange colour for 2017 presence
  legend = FALSE        # Keep legend off for clarity
)
# Overlap = stable areas, non-overlap = change (gain/loss)
```

### Explore rasters on an interactive map for content and comparison (optional)

Lastly, for web maps, rasters need to be in **WGS84** (`EPSG:4326`).

```{r}
# Reproject rasters to WGS84 (EPSG:4326) for Leaflet compatibility if your data is in GDA 94 or other CRS families.
#m2000 <- project(mangrove.2000, "EPSG:4326")   # Convert 2000 raster to lat/long
#m2017 <- project(mangrove.2017, "EPSG:4326")   # Convert 2017 raster to lat/long

m2000 <- mangrove.2000
m2017 <- mangrove.2017

# Note: Leaflet package needs to be loaded for this code to work
 library(leaflet)
 leaflet() |>
   addProviderTiles("CartoDB.Positron") |>   # Add a light basemap for context
   
   addRasterImage(
     m2000,                                  # Add 2000 mangrove layer
     colors = function(x) ifelse(is.na(x), NA, "grey30"), # Grey for presence
     opacity = 0.7,                          # Semi-transparent for overlay
     group = "Mangrove 2000"                 # Layer name for toggle control
   ) |>
   
   addRasterImage(
     m2017,                                  # Add 2017 mangrove layer
     colors = function(x) ifelse(is.na(x), NA, "darkorange"), # Orange for presence
     opacity = 0.7,                          # Same transparency for comparison
     group = "Mangrove 2017"                 # Layer name for toggle control
   ) |>
   
   addLayersControl(
     overlayGroups = c("Mangrove 2000", "Mangrove 2017"), # Toggle between years
     options = layersControlOptions(collapsed = FALSE)    # Keep control panel open
   )
# Toggle layers to visually compare change in mangrove distribution between years
```

### Exercises:

1.  In the Data folder of this workshop, you will find a polygon dataset called WhiteAshForest. Import the data using the function `st_read()` from the sf package, used for shapefile data. Use the help tab to access the function's documentation for correct usage. Note that for shape files, you will need to enter both the name of the folder ("dsn") and the name of the layer ("layer"), and the layer name must match across all the files of the layer.

```{r}
# your code here  
```

2.  Create a plot of the polygon dataset. Note that when plotting polygon data, you must ask the plot function to focus on its geometry using the function `st_geometry()`. What happens to the plot if you don't do this? (Alternatively, if you are comfortable, modify the leaflet plotting code provided above to make an interactive plot.)

```{r}
# your code here  
```

3.  Determine the coordinate reference system (CRS) of the polygon dataset. Is it a geodesic or projected coordinate system? If it's geodesic, project it into an appropriate projected coordinate system for analysis.

```{r}
# your code here 
```

4.  Select a dataset from your own work and import it. Remember to use `rast()` for raster datasets, `st_read()` for shapefiles, and `read.csv()` followed by `st_as_sf()` for point data in table format. Create a plot of your data and determine the CRS.

```{r}
# your code here
```

## Lesson 2: **Extent Of Occurrence --- EOO (Criterion B1)**

For subcriterion B1 , EOO is defined as the **minimum convex polygon** around all occurrences.

It is **not** the same as actual ecosystem area of the species. Instead, it is a standardised spatial metric used to assess vulnerability to broad-scale threatening events.

```{r}
# Create the minimum convex polygon around all ecosystem presence cells in 2017.
# In the current redlistr package, the function name is getEOO().
#EOO <- makeEOO(mangrove.2017)
library(redlistr) # Red List of Ecosystems calculations
EOO <- getEOO(mangrove.2017)


```

The `getEOO()` function outputs a special object containing information about the EOO analysis. It contains "slots" for the name of the assessment unit, the EOO polygon, its area, the units, and the input data used to derive the polygon. The package contains handy default functions for generating summaries and plots of its contents.

```{r}

# Print a summary of the EOO object.
summary(EOO)

# The inbuilt plot function for EOO outputs includes both the input data and the output polygon. 
plot(EOO)

# Extract the EOO area, usually returned in square kilometres.
EOO.area <- getAreaEOO(EOO)

# Alternatively, you can access any slot directly using the @ operator: 
EOO.area <- EOO@EOO

# Print the EOO area.
EOO.area
```

### Exersise:

1.  Get Extent of Occurrence for your own data using the example above.

    ```{r}

    #your code here
    ```

## Lesson 3 : Area of Occupancy --- AOO (Criterion B2)

For subcriterion B2, we will need to calculate the number of grid cells occupied by our species or ecosystem. For species, the grid cell size is 2km x 2km and for ecosystems the cell size is 10km x 10km. The `getAOO()` function uses a random grid search to find the optimal grid with the fewest cells.

Demo

Let's do a demo with our mangrove raster dataset.

```{r Creating AOO grid, fig.width=7, fig.height=7, eval = FALSE}

# getAOO() function 
AOO.grid <- getAOO(mangrove.2017)

```

The function returns a special class of object called an AOOgrid. This object has "slots" for the name of the assessment unit, the best AOO grid, the number of cells in the grid, the parameters used to generate the results, the input data, and the AOO results from any random iterations. Just like EOO, it has some handy built-in methods to summarise and plot your results:

```{r}
# summary
summary(AOO.grid)

# handy plotting function
plot(AOO.grid)
```

You can also pull out individual components of the object using the \@ operator if you want to use them individually (for example, to include in a different plot):

```{r}

# Extract the output grid (for example, to include in a different plot)
grid <- AOO.grid@grid

# Get the number of grid cells as an integer
AOO <- AOO.grid@AOO 
```

You can run some diagnostics on the results to get a sense for whether you ran enough random trials to get the optimal grid.

```{r Getting number of AOO grids, eval = FALSE}

# jplot: we are looking for the last decrease to occur early
jplot(AOO.grid) 

# histogram of values generated with each rep: should look like a bell curve if enough iterations were run. More iterations may produce better results if the left side of the bell curve is missing.
hist(AOO.grid) 

```

The `getAOO()` function has sensible defaults that are readily customisable. Here is a list of arguments that can be modified:

-   cell_size: change to 2000 for species data (default is 10km x 10km)

-   names_from : argument to enter the name of the column containing species or ecosystem labels. Relevant for point or polygon data containing multiple species or ecosystems. If nothing is entered, the function assumes all inputs are part of the same assessment unit.

-   bottom_1pct_rule : turn the bottom 1% rule on or off (defaults to TRUE)

-   percent : change the percent that is cut (defaults to 1%)

-   jitter : turn random iterations on or off, e.g., if you need a quick estimate you can turn it off (defaults to TRUE = on)

-   n_jitter : how many random iterations are desired? (defaults to 35)

My diagnostics indicated that the default number of random iterations may have been insufficient. I want to run my `getAOO` again but with a larger number of iterations to get the best possible results.

```{r}

# increase the number of iterations
AOO.grid <- getAOO(mangrove.2017, n_jitter = 150)

# Run diagnostics again.
# jplot
jplot(AOO.grid)

# histogram of values generated with each rep
hist(AOO.grid)
```

### Multiple units

To demonstrate how to handle datasets containing multiple assessment units, let's look back at our plant species occurrence dataset.

```{r}

# specify the column separating units using names_from
# specify the grid size, usually 2kmx2km for species data
# Turn off the bottom 1% rule -- not relevant for species data
# Force random grid search for all units/species
plant_species_AOO <- getAOO(plant_species, 
                            names_from = "species", 
                            cell_size = 2000, 
                            bottom_1pct_rule = F, 
                            jitter = T)

#install.packages("purrr")
library(purrr)

# the map and walk functions apply a specified operation to each element of a list one by one. 
walk(plant_species_AOO, summary)
map(plant_species_AOO, plot)

```

### Exercises:

1.  Run diagnostics on the AOO of our species point dataset. Does the default number of 35 random iterations seem to converge to the optimal grid?

```{r}
    # your code here


```

2.  Calculate the AOO of our White Ash Forest dataset, then summarise and plot the results.

```{r}
# your code here


```

3.  Calculate AOO for our raster dataset, but turn off the bottom 1% rule. Compare the results for AOO with and without the rule.

```{r}
# your code here

```

4.  Modfiy your code to apply the bottom 1% rule, but dropping the bottom 5%. Compare your results to exercise 3.

```{r}
# your code here

```

5.  Apply the AOO function to your own data and run diagnostics. Do your results agree with any previous assessments? Why or why not?

```{r}
# your code here

```

## Lesson 4 : Calculating rates of change (45 minutes)

Criterion A of the Red List of Ecosystems and Criteria A-C of the Red List of Threatened Species all deal with calculating rates of change, often with very few observations to support the calculations. In many cases, all we have are two measurements of area or population from several years apart. In this section we will cover some basic strategies for making estimates of change over time.

### I have two time slices

For ecosystems or species ranges, we first calculate area of our measurements through time.

Area is calculated by counting raster cells classified as the ecosystem and multiplying by cell size.

Results are returned in **km²**.

```{r Calculate area of rasters}
a.2000 <- getArea(mangrove.2000)
a.2017 <- getArea(mangrove.2017)

# Combine multiple rasters and calculate their area in one step.
# Here we pass both 2000 and 2017 layers together.

a <- getArea(c(mangrove.2000, mangrove.2017))
```

Typically we will want to know the difference in the areas or populations we calculated in the previous step.

```{r}
# The getAreaChange function can accept rasters or polygons (recommended)

area.change <- getAreaChange(mangrove.2000, mangrove.2017)

# It can also accept the area tables calculated in the previous step, but the column containing an ecosystem identity value must be specified. 

area.change <- getAreaChange(a.2000, a.2017, value) 
```

In the Red List of Ecosystems, two methods are suggested to determine the rate of decline of an ecosystem, each of which assumes a different functional form of the decline ([IUCN, 2024](https://portals.iucn.org/library/sites/library/files/documents/2024-021-En.pdf)). The proportional rate of decline (PRD) is a fraction of the previous year's remaining area, while the absolute rate of decline (ARD) is a constant fraction of the area of the ecosystem at the beginning of the decline ([IUCN, 2024](https://portals.iucn.org/library/sites/library/files/documents/2024-021-En.pdf)). These rates of decline allow the use of two or more data points to extrapolate to the full 50 year timeframe required in an assessment.

The annual rate of change (ARC) uses a compound interest law to determine the instantaneous rate of change ([Puyravaud 2003](https://www.sciencedirect.com/science/article/pii/S0378112702003353)).

To estimate the rate of changing using these methods, we use the `getDeclineStats` function.

```{r Using getDeclineStats}
decline.stats <- getDeclineStats(a.2000$area, a.2017$area, 2000, 2017, 
                                 methods = c('ARD', 'PRD', 'ARC'))
decline.stats
```

Extrapolate to appropriate time frame.

```{r Estimating future area}
extrapolated.area <- extrapolateEstimate(a.2000$area, year.t1 = 2000, 
                                        ARD = decline.stats$ARD, 
                                        PRD = decline.stats$PRD, 
                                        ARC = decline.stats$ARC, 
                                        nYears = 50)
extrapolated.area

```

### Exercises:

1.  Repeat the calculations of area/rates of change on an example polygon dataset.

    ```{r}

    #your code here
    ```

2.  Extrapolate -- are three decline modes in agreement?

    ```{r}

    #your code here
    ```

3.  Calculate the estimated rate of change in the population of XXX using PRD and extrapolate to XX years.

    ```{r}

    #your code here
    ```

4.  Use a dataset of your own to estimate the rate of change and the percent change in area/population over a time frame that is relevant to your ecosystem or species.

    ```{r}

    #your code here
    ```

## Whats's next?

**I have a time series (3 or more measurements)**

-   Areas

-   Trend fitting.

-   Extrapolation

### Thank you for your time :)
