---
title: "Workshop: Introduction to 'redlistr'" # Title of the workshop page
author: "Abhimanyu raj Singh"
format: html
editor: visual # Prefer Quarto's visual editor mode in supported editors
---

## Overview -Lesson plan

A beginner workshop for using the `{redlistr}` package in **R** to calculate key spatial metrics used in the **IUCN Red List of Ecosystems** workflow from ecosystem distribution rasters, [*shapefiles and datasheets.*]{.underline}

Lesson 0: Introduction to R, RStudio, and Quarto

Lesson 1: Setting up the environment and loading packages

Lesson 2: Getting the data

Lesson 3: Area calculation

Lesson 4: Extent of Occurrence (EOO)

Lesson 3: Area of Occupancy (AOO)

Lesson 4: Rate of change

## Lesson 0: Introduction to R, Rstudio and Quarto

Before starting, install:

- R (language)
- RStudio (IDE - Integrated Development Environment)
- Quarto (optional, but useful for workshop documents like this one)

[You will also need a folder for your workshop data. In this workshop we will use one main workspace folder and keep the processing files inside it.]{.underline}

``` text
Workshop_redlistr/
├── Workshop.qmd
```

## Lesson 1: Setting up environment and loading packages

### 1.1 Install Packages

Run this once on your computer if the packages are not already installed:

```{r}
#| eval: false
# Install the packages needed for this workshop.
# Run this only once on your computer, not every session.
install.packages(c(
  "terra",           # raster handling
  #"leaflet",         # interactive maps
  "redlistr"         # Red List of Ecosystems helper functions
))
```

### 1.2 Load Packages

Run this in each session:

```{r Loading packages, message=FALSE}
# Load packages
library(terra) # work with raster data
#library(leaflet) # build interactive maps
library(redlistr) # Red List of Ecosystems calculations
```

### 1.3 Set Your Workspace

### The easiest habit for beginners is to create one object called `Workspace` and build all file paths from it.

```{r}
# We will keep your default directory for workshop use today and assign it as our workspace
Workspace <- getwd()

# Create the local workspace folder if it does not exist yet
dir.create(Workspace, recursive = TRUE, showWarnings = FALSE)
```

## Lesson 2: Get data, run checks and explore

In this step, we first get the data, then visualise them and check if our data matches the metadata.

Here we will use the mangrove rasters bundled inside `{redlistr}`. Our data-set is based on Mangrooves distributions around French island VIC.

```{r Loading our example distributions}
# Read the example mangrove raster for year 2000 from the redlistr package.
mangrove.2000 <- rast(
  system.file("extdata", "example_distribution_2000.tif", package = "redlistr")
)

# Read the example mangrove raster for year 2017 from the redlistr package.
mangrove.2017 <- rast(
  system.file("extdata", "example_distribution_2017.tif", package = "redlistr")
)
```

- Here, `system.file()` points to files bundled inside the package.

- **When using your own data, replace this with a full file path, for example:** **`rast("C:/mydata/mangroves_2000.tif")`**

### 2.1 Visualise both years together

This first check is important. Before doing any calculations, inspect what the rasters looks like.

`add = TRUE` overlays the 2017 map on top of the 2000 map.

- **Grey** = 2000

- **Orange** = 2017

```{r Plotting the two rasters, fig.show='hold', fig.width=7, fig.height=7, fig.alt='Overlay of two mangrove distribution maps'}
plot(
  mangrove.2000,        # Base layer: mangrove distribution in year 2000
  col = "grey30",       # Grey colour for 2000 presence
  legend = FALSE,       # Remove legend for cleaner overlay
  main = "Mangrove Distribution"  # Plot title
)

plot(
  mangrove.2017,        # Overlay layer: mangrove distribution in 2017
  add = TRUE,           # Add on top of existing plot (do not create new plot)
  col = "darkorange",   # Orange colour for 2017 presence
  legend = FALSE        # Keep legend off for clarity
)
# Overlap = stable areas, non-overlap = change (gain/loss)
```

### 2.2 Data preparation checks for our calculations: Check CRS of **both** rasters

Second step is also **critical before any calculations**.

The `{redlistr}` functions require distances in **metres**, not degrees.

Your rasters must:

- be in a **projected CRS** rather than geographic longitude/latitude

- using the **same CRS**

Before doing any calculations, always check the Coordinate Reference System (CRS).

**Definition:** CRS (Coordinate Reference System) defines how spatial data are positioned on the Earth (e.g., latitude/longitude or projected metres).

**Definition:** Projected CRS represents data in metres, which is required for accurate area and distance calculations.

```{r}
# Print CRS of 2000 raster
crs(mangrove.2000)

# Print CRS of 2017 raster
crs(mangrove.2017)

# Example only: choose a projected CRS suitable for your study area
# mangrove.2000 <- project(mangrove.2000, "EPSG:32754")
# mangrove.2017 <- project(mangrove.2017, "EPSG:32754")
```

#### Check if data are in longitude/latitude (degrees)

```{r}
# TRUE = geographic (degrees), FALSE = projected (metres)
is.lonlat(mangrove.2000)
is.lonlat(mangrove.2017)

```

If `is.lonlat()` returns TRUE, your data are in degrees (longitude/latitude). You must reproject them before continuing.

#### Compare CRS between layers

```{r}
# Check if both rasters use the same CRS
crs(mangrove.2000) == crs(mangrove.2017)
```

### 2.3 Explore rasters on the interactive map for content and comparison

Lastly, for web maps, rasters need to be in **WGS84** (`EPSG:4326`).

```{r}
# Reproject rasters to WGS84 (EPSG:4326) for Leaflet compatibility if you data is in GDA 94 or others CRS family.
#m2000 <- project(mangrove.2000, "EPSG:4326")   # Convert 2000 raster to lat/long
#m2017 <- project(mangrove.2017, "EPSG:4326")   # Convert 2017 raster to lat/long

leaflet() |>
  addProviderTiles("CartoDB.Positron") |>   # Add a light basemap for context
  
  addRasterImage(
    m2000,                                  # Add 2000 mangrove layer
    colors = function(x) ifelse(is.na(x), NA, "grey30"), # Grey for presence
    opacity = 0.7,                          # Semi-transparent for overlay
    group = "Mangrove 2000"                 # Layer name for toggle control
  ) |>
  
  addRasterImage(
    m2017,                                  # Add 2017 mangrove layer
    colors = function(x) ifelse(is.na(x), NA, "darkorange"), # Orange for presence
    opacity = 0.7,                          # Same transparency for comparison
    group = "Mangrove 2017"                 # Layer name for toggle control
  ) |>
  
  addLayersControl(
    overlayGroups = c("Mangrove 2000", "Mangrove 2017"), # Toggle between years
    options = layersControlOptions(collapsed = FALSE)    # Keep control panel open
  )
# Toggle layers to visually compare change in mangrove distribution between years
```

##### For Alternative formats

See vignette [Ecosystem distribution data in raster and vector format](redlistr-raster-vector.html) for instruction on how to read ecosystem distribution data in other formats like shapefiles.

### Exersises:

1.  Get your own raster data

    ```{r}

    #your code here
    ```

2.  Get your own vector data

    ```{r}

    #your code here
    ```

3.  Check CRS of you data and compare between layers

    ```{r}

    #your code here
    ```

4.  Check if you data is in meters

    ```{r}

    #your code here
    ```

## Lesson 4: **Calculate extent of occurrence --- EOO (Criterion B1)**

For subcriterion B1 , EOO is defined as the **minimum convex polygon** around all occurrences.

It is **not** the same as actual ecosystem area of the species. Instead, it is a standardised spatial metric used to assess vulnerability to broad-scale threatening events.

```{r}
# Create the minimum convex polygon around all ecosystem presence cells in 2017.
# In the current redlistr package, the function name is makeEOO().
EOO.polygon <- makeEOO(mangrove.2017)

# Print a summary of the EOO object.
summary(EOO.polygon)

# Plot the 2017 raster first so the EOO polygon has spatial context.
plot(mangrove.2017, col = "lightgrey", legend = FALSE, main = "EOO for mangrove distribution in 2017")

# Add the EOO polygon on top of the raster.
plot(EOO.polygon, add = TRUE, border = "red", lwd = 2, col = NA)

# Extract the EOO area, usually returned in square kilometres.
EOO.area <- getAreaEOO(EOO.polygon)

# Print the EOO area.
EOO.area
```

### Exersise:

1.  Get Extent of Occurrence for your own data using the example above.

    ```{r}

    #your code here
    ```

**⚠️ Important technical notes**

- EOO and AOO should be calculated using standard methods only. Non-standard approaches cannot be compared directly against IUCN thresholds.

**CRS mismatch pitfalls**

- `{terra}` → projected CRS (good for area)

- `{leaflet}` → requires **EPSG:4326**

- So we maintain **two versions**:

  - projected → analysis

  - WGS84 → visualisation

## [6. Placeholder for later workshop steps]{.underline}

[We are focusing on **EOO only** in this version of the workshop.]{.underline}

[You can add the following sections later if needed:]{.underline}

- [**AOO (Area of Occupancy)**]{.underline}

- [**Rate of change**]{.underline}

- [**Time series assessment**]{.underline}

## Lesson 3: **Ecosystem** Area Calculation **(Criterion A)**

For Criterion A, we will question: **How big is the ecosystem at each time point?**

Area is calculated by counting raster cells classified as the ecosystem and multiplying by cell size.

Results are returned in **km²**.

```{r}
# Calculate area for each raster value in the 2000 layer.
# This is useful for binary rasters because it separates absence and presence values.
a.2000 <- getArea(mangrove.2000, byValue = TRUE)

# Calculate area for each raster value in the 2017 layer.
a.2017 <- getArea(mangrove.2017, byValue = TRUE)

# Print the full area tables.
#a.2000
#a.2017

# Area = total area (km²) for each class
# Combine multiple rasters and calculate their area in one step.
# Here we pass both 2000 and 2017 layers together.
a.both <- getArea(c(mangrove.2000, mangrove.2017), byValue = TRUE)

# Print the combined result.
a.both
#if needed a combined area
#sum(a$area)
```

- `getArea()` also works on raster stacks with multiple ecosystem layers. This is useful when scaling up to many ecosystems, regions, or time points.

### Exersise:

1.  Get Area for your