---
title: "Workshop: Introduction to 'redlistr'" # Title shown at the top of the rendered workshop page
editor: visual # Prefer Quarto's visual editor mode in supported editors
---

## Overview

A beginner workshop for using the `{redlistr}` package in **R** to calculate key spatial metrics used in the **IUCN Red List of Ecosystems** workflow from ecosystem distribution rasters.

**Lesson 1: Setup**

a.  Introduction to R, RStudio, and Quarto

b.  Setting up the environment

c.  Importing data

d.  Visualising data

**Lesson 2: Extent of Occurrence (EOO)**

**Lesson 3: Area of Occupancy (AOO)**

**Lesson 4: Calculating Area and Rate of change**

## Lesson 1. Setup

### a. Introduction to R, Rstudio and Quarto

Before starting, install:

-   R
-   RStudio
-   Quarto (optional, but useful for workshop documents like this one)

We want to keep all of our files in one folder. Download a copy of the [GitHub repository](https://github.com/red-list-ecosystem/redlistr_intro_workshop.git) (use the green Code button at the top right), and save it in a convenient place on your computer. The repository will be your main workspace for this workshop. Open the workshop in Rstudio using the `redlistr_intro_workshop.Rproj` file. This will automatically set your working directory to the repository folder.

### b. Setting up environment

#### Install Packages

Run this once on your computer if the packages are not already installed:

```{r}
#| eval: false
# Install the packages needed for this workshop.
# Run this only once on your computer, not every session.
install.packages(c(
  "sf",              # shapefile handling
  "terra",           # raster handling
  "leaflet",         # interactive maps
  "redlistr"         # Red List of Ecosystems helper functions
))

### TODO: If we don't publish to CRAN, we will need: 
install.packages("remotes")
remotes::install_github("red-list-ecosystem/redlistr_intro_workshop")
# if at this point you are prompted to update packages, enter 1 to update ALL and click "yes" in the popup window.

```

#### Load Packages

Run this each session:

```{r Loading packages, message=FALSE}
# Load packages
library(sf) # work with point, line, and polygon data
library(terra) # work with raster data
library(leaflet) # build interactive maps
library(redlistr) # Red List of Ecosystems calculations
```

### c. Importing Data

Here we use the mangrove rasters bundled inside `{redlistr}`.

```{r Loading our example distributions}
# Read the example mangrove raster for year 2000 from the redlistr package.
mangrove.2000 <- rast(
  system.file("extdata", "example_distribution_2000.tif", package = "redlistr")
)

# Read the example mangrove raster for year 2017 from the redlistr package.
mangrove.2017 <- rast(
  system.file("extdata", "example_distribution_2017.tif", package = "redlistr")
)
```

`system.file()` points to files bundled inside the package.

`rast()` specifies that the file is to be read in as a raster file.

When using your own raster data, replace this with a full file path, for example:\`rast("C:/mydata/mangroves_2000.tif")\`

#### Alternative formats

See vignette [Ecosystem distribution data in raster and vector format](redlistr-raster-vector.html) for instruction on how to read ecosystem distribution data in other formats.

### d. Visualising data

This first check is important for beginners. Before doing any calculations, inspect what the rasters look like.

```{r Plotting the two rasters, fig.show='hold', fig.width=7, fig.height=7, fig.alt='Overlay of two mangrove distribution maps'}
plot(
  mangrove.2000,        # Base layer: mangrove distribution in year 2000
  col = "grey30",       # Grey colour for 2000 presence
  legend = FALSE,       # Remove legend for cleaner overlay
  main = "Mangrove Distribution"  # Plot title
)

plot(
  mangrove.2017,        # Overlay layer: mangrove distribution in 2017
  add = TRUE,           # Add on top of existing plot (do not create new plot)
  col = "darkorange",   # Orange colour for 2017 presence
  legend = FALSE        # Keep legend off for clarity
)
# Overlap = stable areas, non-overlap = change (gain/loss)
```

\`add = TRUE\` overlays the 2017 map on top of the 2000 map.

-   **Grey** = 2000

-   **Orange** = 2017

Therefore, areas that show up in grey are those where mangrove distribution has been lost between 2000 and 2017.

#### Explore rasters on an interactive map for content and comparison (optional)

For web maps, rasters need to be in **WGS84** (`EPSG:4326`).

```{r}
# Reproject rasters to WGS84 (EPSG:4326) for Leaflet compatibility
m2000 <- project(mangrove.2000, "EPSG:4326")   # Convert 2000 raster to lat/long
m2017 <- project(mangrove.2017, "EPSG:4326")   # Convert 2017 raster to lat/long

leaflet() |>
  addProviderTiles("CartoDB.Positron") |>   # Add a light basemap for context
  
  addRasterImage(
    m2000,                                  # Add 2000 mangrove layer
    colors = function(x) ifelse(is.na(x), NA, "grey30"), # Grey for presence
    opacity = 0.7,                          # Semi-transparent for overlay
    group = "Mangrove 2000"                 # Layer name for toggle control
  ) |>
  
  addRasterImage(
    m2017,                                  # Add 2017 mangrove layer
    colors = function(x) ifelse(is.na(x), NA, "darkorange"), # Orange for presence
    opacity = 0.7,                          # Same transparency for comparison
    group = "Mangrove 2017"                 # Layer name for toggle control
  ) |>
  
  addLayersControl(
    overlayGroups = c("Mangrove 2000", "Mangrove 2017"), # Toggle between years
    options = layersControlOptions(collapsed = FALSE)    # Keep control panel open
  )
# Toggle layers to visually compare change in mangrove distribution between years
```

#### Data preparation checks for our calculations

It is critical to perform two checks **before any calculations**.

The {redlistr} functions require distances in **metres**, not degrees.

Your data must:

-   be in a **projected CRS** rather than geographic longitude/latitude

-   use the **same CRS**

Before doing any calculations, always check the Coordinate Reference System (CRS). The package will check this internally and throw an error if datasets are provided with geographic coordinates.

**Definition:** CRS (Coordinate Reference System) defines how spatial data are positioned on the Earth (e.g., latitude/longitude -- geodesic coordinate systems that model a round surface or projected coordinate systems where coordinates are projected onto a flat surface).

**Definition:** Projected CRS represents data in metres (or feet), which is required for accurate area and distance calculations, as degrees can represent different distances depending on latitude.

Note: For Red List assessments, it is best to choose a CRS that compromises between accurate areas, distances, and bearings. For smaller extents, the exact CRS will not affect the results, but for analyses covering larger, regional areas, we might like to test if different projections yield different results.

#### Check CRS of **both** rasters

```{r}
# Print CRS of 2000 raster
st_crs(mangrove.2000)

# Print CRS of 2017 raster
st_crs(mangrove.2017)
```

#### Check if data are in longitude/latitude (degrees)

```{r}
# TRUE = geographic (degrees), FALSE = projected (metres)
is.lonlat(mangrove.2000)
is.lonlat(mangrove.2017)
```

If `is.lonlat()` returns TRUE, your data are in degrees (longitude/latitude).

You must reproject them before continuing. We will not cover selecting an appropriate CRS in this workshop, but external resources for this step are available.

```{r}
# Example only: choose a projected CRS suitable for your study area
# mangrove.2000 <- project(mangrove.2000, "EPSG:32754")
# mangrove.2017 <- project(mangrove.2017, "EPSG:32754")
```

#### Compare CRS between layers

```{r}
# Check if both rasters use the same CRS
crs(mangrove.2000) == crs(mangrove.2017)
```

### Exercises:

1.  In the Data folder of this workshop, you will find a polygon dataset called XXX. Import the data using the function `st_read()` from the sf package, used for polygon and point data. Use the help tab to access the function's documentation for correct usage.

```{r}
# your code here  
```

2.  Create a plot of the dataset. Note that when plotting polygon data, you must ask the plot function to focus on its geometry using the function `st_geometry()`. What happens to the plot if you don't do this?

```{r}
# your code here  
```

3.  Determine the coordinate reference system (CRS) of the polygon dataset. Is it a geodesic or projected coordinate system? If it's geodesic, project it into an appropriate projected coordinate system for analysis.

```{r}
# your code here 
```

4.  Select a dataset from your own work and import it. Create a plot and determine the CRS.

```{r}
# your code here
```

## Lesson 2: **Calculate extent of occurrence --- EOO (Criterion B1)**

EOO is the **minimum convex polygon** around all occurrences.

It is **not** the same as actual ecosystem area.

Instead, it is a standardised spatial metric used to assess vulnerability to broad-scale threatening events.

```{r}
# Create the minimum convex polygon around all ecosystem presence cells in 2017.
# In the current redlistr package, the function name is getEOO().
EOO <- getEOO(mangrove.2017)

```

The `getEOO()` function outputs a special object containing information about the EOO analysis. It contains "slots" for the EOO polygon, its area, the units, and the input data used to derive the polygon. The package contains handy default functions for generating summaries and plots of its special functions.

```{r}

# Print a summary of the EOO object.
summary(EOO)

# The inbuilt plot function for EOO outputs includes both the input data and the output polygon. 
plot(EOO)

# Extract the EOO area, usually returned in square kilometres.
EOO.area <- getAreaEOO(EOO)

# Alternatively, you can access any slot directly using the @ operator: 
EOO.area <- EOO@area

# Print the EOO area.
EOO.area

# You can always use the str() function to get an overview of the contents of an object, for instance if you don't know what the slots are named:
str(EOO)
```

**⚠️ Important technical notes**

EOO and AOO should be calculated using standard methods only.

Non-standard approaches cannot be compared directly against IUCN thresholds.

**CRS mismatch pitfalls**

-   {terra} → projected CRS (good for area)

-   {leaflet} → requires **EPSG:4326**

-   So you maintain **two versions**:

    -   projected → analysis

    -   WGS84 → visualisation

### Summary

```         
Raw raster → CRS check → Projection (UTM) → Area (Criterion A)
                                          → EOO (Criterion B)

                           ↓
                   Reproject to WGS84
                           ↓
                  Interactive visualisation (Leaflet)
```

### Exercises:

1.  Generate the EOO object for mangrove.2000.

```{r}
# your code here  
```

2.  View the summary and plot of the EOO object. What is the EOO area? Did it increase or decrease between 2000 and 2017?

```{r}
# your code here  
```

3.  Repeat exercises 1 and 2 for the polygon dataset we loaded earlier.

```{r}
# your code here 
```

4. Repeat exercises 1 and 2 for a dataset of your own. 

```{r}
# your code here
```

